Short and Mechanized Logical Relation for Dependent Type Theory

### 2024-02-22 paper/paper.tex
Revisions to Sec 4-6

\scw{Need to explain the notations in the text. What are these operations?
What do readers need to understand about them? Help me understand the figure}
\scw{As this is a tutorial paper, you'll need to explain more about how de Bruijn
indices work.}
% We omit most of the
% definitions of renaming and substitution and only show the definition
% of a few representative cases of substitution.
\begin{figure}[ht]
  \begin{equation*}
    \begin{split}
      \begin{array}{lll}
        % \mathit{IdentityRen} \\
        % [[id i]] & := & [[i]] \\ \\
        \mathit{IdentityTm} \\
        [[idtm n]] & := & [[var n]] \\ \\
        % \mathit{ConsRen ([[xi .: n]])} \\
        % ([[xi .: n]])([[0]]) & := & [[n]] \\
        % ([[xi .: n]])([[Suc i]]) & := & [[xi i]] \\ \\
        \mathit{ConsSubst ([[rho .: a]])} \\
        ([[rho .: a]])([[0]]) & := & [[a]] \\
        ([[rho .: a]])([[Suc n]]) & := & [[rho n]] \\ \\
        \mathit{Curried\ Add} \\
        [[up n m]] & := & \ottkw{v}_{[[n]] + [[m]]}
      \end{array}
    \end{split}
    \qquad \qquad
    \begin{split}
      \begin{array}{lll}
        % \mathit{UpRen} \\
        % [[( up xi ) 0]] & := & [[0]] \\
        % [[( up xi ) Suc i]] & := & [[ Suc ren xi i ]] \\ \\
        % \mathit{Renaming ([[a < xi >]])} \\
        % [[(Pi A B) < xi >]] & := & [[Pi A < xi > B < up xi >]] \\
        % [[(a b) < xi >]] & := & [[a < xi > (# b < xi > #)]] \\
        % \ldots \\ \\
        \mathit{Lifting} \\
        [[( up rho ) 0]] & := & [[var 0]] \\
        [[( up rho ) Suc n]] & := & [[ rho n < up 1 > ]] \\ \\
        \mathit{Substitution ([[ a { rho }  ]])} \\
        [[var n { rho }  ]] & := &  [[rho n]] \\
        [[(Pi A B) { rho }]] & := & [[Pi A { rho } B { up rho }]] \\
        [[(a b) { rho }]] & := & [[a { rho } (# b { rho } #)]] \\
        [[(\ a) { rho }]] & := & [[\ (a { up rho })]] \\
        \ldots
      \end{array}
    \end{split}
  \end{equation*}
  \caption{Auxiliary Functions over Syntax}
  \label{fig:auxdef}
\end{figure}

@@

\section{Semantic Typing and Consistency}
\label{sec:logrelproof}
\begin{figure}[h]
\[
\begin{array}{lcl}

@@

The necessity of semantic typing as an extra layer of definition on
top of the logic relation can be understood in simply typed
languages~\citep{skorstengaard2019introduction, harpertait,
  pierce2002types}. In our setting, attempting to show that $[[empty
|- a : A]]$ implies $[[InterpR i A S]]$ and $[[a in S]]$ will fail in
\rref{T-Abs}, where induction hypothesis is not helpful since the body
of the lambda term is typed under a non-empty context. Through
the definition of semantic typing, we can state a strengthened
property that is actually provable.

@@

The $[[rho |= G]]$ judgment defines the semantic well-formedness of
a substitution $[[rho]]$ with respect to a context
$[[G]]$. Intuitively, for variable $[[i]]$ in the context (thus the
restriction $[[i < | G |]]$), $[[rho i]]$ is a term that inhabits all
possible interpretations of the type $[[G i]]$ from the
context. However, since we are in a dependently typed system, there
may be variables in $[[G i]]$, too. The type $[[(G i < up Suc i > ) {
  rho }]]$ applies $[[rho]]$ to replace those variables. The shifting
operation $[[up Suc i]]$ can be viewed as an artifact of the de Bruijn
representation to lift the type $[[G i]]$ to the right scope before
substitution can be carried out.

In the
definition of $[[|= G]]$, we use the notation $[[drop i G]]$ to denote
the typing context obtained by dropping the last $[[i]]$ elements of
$[[G]]$. When $[[G]]$ has less than $[[i]]$ elements, $[[drop i G]]$
returns the empty list.

The following lemma makes the statement $[[G |= A : Set i]]$ easier to
work with.
\begin{lemma}[Set Inv\footnote{\dotv{soundness.v}{SemWt\_Univ}}\scw{spell out the whole word}]
  \label{lemma:setinv}
  The following two statements are equivalent:
  \begin{itemize}
  \item $[[G |= A : Set i]]$
  \item $\forall$ $[[rho]]$, if $[[rho |= G]]$, then there exists
    $[[S]]$ such that $[[InterpR i (A {rho}) S]]$
  \end{itemize}
\end{lemma}
\begin{proof}
  The forward direction is immediate by
  Lemma~\ref{lemma:interpinv}. We now consider the backward direction
  and show that $[[G |= A : Set i]]$ given the second bullet.

@@

The semantic context well-formedness judgment ($[[|= G]]$), unlike its syntactic
counterpart $[[|- G]]$, is defined through a $\forall$ quantified statement rather
than inductively over the context. \scw{why is that the case?} It is easy to recover the same
structural rules:
\begin{lemma}[Semantic context well-formedness cons]
  \label{lemma:semwffcons}
  If $[[|= G]]$ and $[[G |= A : Set i]]$, then $[[|= G ++ A]]$.
\end{lemma}
\scw{point out that this lemma is the semantic analogue to Ctx-Cons. What about
Ctx-empty?}
\begin{proof}
  By the definition of semantic context well-formedness, the goal is
  to show that given $[[i < | G ++ A |]] = [[Suc |G|]]$, $[[drop Suc i
  (G ++ A) |= (G ++ A) i : Set j]]$. The statement can be easily
  proven by case analysis on whether $[[i]]$ is zero.
\end{proof}

Likewise, the semantic well-formendness judgment for substitutions
satisfies similar structural rules.
\begin{lemma}[Well-formed $[[rho]]$ cons\footnote{\dotv{soundness.v}{$\gamma$\_ok\_cons}}]
  If $[[InterpR i A S]]$, $[[a in S]]$, and $[[rho |= G]]$, then
  $[[rho .: a |= G ++ A]]$.
\end{lemma}
\begin{proof}
  Start by unfolding the definition of $[[rho .: a |= G ++ A]]$ and
  performing a case analysis similar to the proof of
  Lemma~\ref{lemma:semwffcons}.  The case where the number is $[[0]]$
  requires Lemma~\ref{lemma:logreldeterhet} to finish the proof.
\end{proof}

Next, we show some non-trivial cases of the fundamental theorem as
top-level lemmas and leave the remaining cases as exercises for the
reader.\scw{I dislike leaving ``exercises to the reader''}

First, we formulate the definition of valid renamings and prove that
semantic typing satisfies renaming so we can weaken the context when
reasoning about the variable case of the fundamental lemma
(Lemma~\ref{lemma:stvar}). Intuitively, given a valuation $[[rho |= G
++ D]]$, it is easy to show that we can extract some valuation
$[[rho0]]$ such that $[[rho0 |= G]]$, where $[[rho0]]$ is obtained by
``truncating'' $[[rho]]$. As a result, if we know that $[[G |= a :
A]]$, then we can conclude that $[[G ++ D |= a0 : A0 ]]$, where
$[[a0]]$ and $[[A0]]$ are obtained by shifting $[[a]]$ and $[[A]]$
after weakening the context; this implication holds because $[[rho |=
G ++ D]]$ induces a context $[[rho0]]$ such that $[[rho0 |= G]]$ so we
can make use of the premise $[[G |= a : A]]$ to derive what we need
for the conclusion. We recommend the readers to skip the proofs of
Lemmas~\ref{lemma:validtruncate} through \ref{lemma:semrenaming}
during the first read as long as they have an intuitive understanding
of what the renaming property is meant to capture.

\scw{Make this a definition? Would it be useful to create notation
  for the relation, such as $[[xi]]:[[G]]\Rightarrow[[D]]$? }
We say that $[[xi]]$ is
valid from the context $[[G]]$ to the context $[[D]]$ if the
following condition holds.
\[ \forall i \text{, if } i < | [[G]] |\text{, then }[[ ren xi i < |D| ]] \text{ and }
  [[D ren xi
  i < up Suc ren xi i > = G i < up Suc i > < xi >]] \]

\begin{lemma}[Truncate is valid]
  \label{lemma:validtruncate}
  If $[[i < |G|]]$, then $[[up i]]$ is a valid renaming from $[[drop i
  G]]$ to $[[G]]$.
\end{lemma}
\begin{proof}
  By the definition of a valid renaming, we must show that given $[[j
  < |drop i G|]] = [[|G|]] - [[i]]$, then the following conditions hold:
  \begin{itemize}
  \item $[[ren up i j]] = i + j  < [[| G | ]]$
  \item $[[G ren up i j < up Suc ren up i j > = drop i G j < up Suc j > < up i >]]$
  \end{itemize}
  The first bullet point is immediate from the fact that $[[j]] <
  [[|G|]] - [[i]]$. The second bullet follows from unfolding the
  definitions, and the fact that $[[drop i G j]] = [[G]](i + j)$ when
  $[[i]] + [[j]] < [[|G|]]$.
\end{proof}

\begin{lemma}[Renaming for $[[rho |= G]]$]
  \label{lemma:renamingval}
  If $[[rho |= D]]$ and $[[xi]]$ is a valid renaming from $[[G]]$ to
  $[[D]]$, then we have $[[rho0 |= G]]$ where $[[rho0 i = rho ren xi i]]$.
\scw{could you use $[[rho]]\circ[[xi]]$ for $[[rho0]]$?}
\end{lemma}
\begin{proof}
  Unfolding the definition of $[[rho0 |= G]]$, the goal is to show
  that for all $i,j,$ and $S$, if $[[i < |G|]]$ and $[[InterpR j (G i
  < up Suc i > ) { rho0 } S ]] \text{, then } [[rho0 i]] = [[rho ren xi i in S]]$.

  By the definition of $[[rho0]]$ and the validity of $[[xi]]$, we
have $[[G i < up Suc i > { rho0 }]] = [[G i < up Suc i > < xi > { rho
} ]] = [[D ren xi i < up Suc ren xi i > { rho }]]$. From $[[rho |=
D]]$ and $[[ren xi i < | D |]]$, we have $[[InterpR j D ren xi i < up Suc ren xi i > { rho }
S]]$ and $[[rho ren xi i in S]]$.
\end{proof}

\begin{lemma}[Renaming for $[[G |= a : A]]$]
  \label{lemma:semrenaming}
  If $[[G |= a : A]]$ and $[[xi]]$ is a valid renaming from $[[G]]$ to
  $[[D]]$, then $[[D |= a < xi > : A < xi > ]]$.
\end{lemma}
\begin{proof}
  Immediate from the definition of semantic typing and Lemma~\ref{lemma:renamingval}.
\end{proof}

\scw{Need a transition here that you are starting to explain the semantic typing
rules.}

\begin{lemma}[ST-Var]
  \label{lemma:stvar}
  If $[[|= G]]$ and $[[i < |G|]]$, then $[[G |= var i : G i < up Suc i  >]]$.
\end{lemma}
\begin{proof}
  Suppose $[[rho |= G]]$. By the definition of semantic typing, we
need to show that there exists some $[[j]]$ and $[[S]]$ such that
  \begin{itemize}
  \item $[[InterpR j G i < up Suc i > { rho } S]]$
  \item $[[rho i in S]]$
  \end{itemize}
  From the definition of $[[|= G]]$, we know that there exists some
  $[[j]]$ such that $[[drop Suc i G |= G i : Set j]]$. By
  Lemma~\ref{lemma:validtruncate}, we know that $[[up Suc i]]$ is a
  valid renaming from $[[drop Suc i G]]$ to $[[G]]$. By
  Lemma~\ref{lemma:semrenaming}, we deduce $[[G |= G i < up Suc
  i > : Set j]]$. By the identity from Lemma~\ref{lemma:setinv}, we
  know that $[[InterpR j G i < up Suc i > { rho } S]]$. It now
  suffices to show $[[rho i in S]]$, but that is immediate from the
  definition of $[[rho |= G]]$.
\end{proof}

@@

  By Lemma~\ref{lemma:setinv} and the premise $[[G |= Pi A B : Set
  i]]$, there exists some set $[[S]]$ such that
  $[[InterpR i Pi A {rho} B {up rho} S]]$. It now suffices to show that
  $[[\A{rho} b{up rho} in S
  ]]$. By Lemma~\ref{lemma:piinvalt}, there exists some $[[S0]]$ such
  that all following conditions hold:
  \begin{itemize}
  \item $[[InterpR i A{rho} S0]]$
  \item $[[forall a, (# a in S0 implies (# exists S1 , InterpR i B
    {rho .: a}
    S1 #) #)]]$
  \item $[[S = { b | forall a, (# a in S0 , forall
      S1, (# InterpR i B {rho .: a} S1,  b a in S1 #) #) }]]$
  \end{itemize}
  To show that $[[\A{rho} b{up rho} in S]]$, we need to prove
  that given $[[a in S0]]$,
  $[[Interp I i B {rho .: a} S1]]$, we have  $[[( \A{rho} b{up rho} )
  a in S1]]$.
  By Lemma~\ref{lemma:logrelbackclos}, the set $[[S1]]$ is closed
  under expansion. By Lemma~\ref{lemma:parabscbn}, since $[[( \A{rho} b{up rho} )
  a => b {up rho} {a}]] = [[b {rho .: a}]]$, it suffices to show that
  $[[b {rho .: a} in S1]]$, which is immediate from $[[G ++ A |= b :
  B]]$ and the fact that the logical relation is deterministic and
  cumulative (Lemma~\ref{lemma:logreldeterhet}).
\end{proof}

@@

\begin{theorem}[The Fundamental Theorem\footnote{\dotv{soundness.v}{soundness}}]
  \label{theorem:soundness}\leavevmode
  \begin{itemize}
  \item If $[[G |- a : A]]$, then $[[G |= a : A]]$.
  \item If $[[|- G]]$, then $[[|= G]]$.
  \end{itemize}
\end{theorem}
\begin{proof}
  Proof by mutual induction over the derivation of $[[G |- a :
  A]]$ and $[[|- G]]$.   The cases related to context well-formedness immediately follows
  from Lemma~\ref{lemma:semwffcons}.
Lemmas~\ref{lemma:stvar},~\ref{lemma:stset},~\ref{lemma:stpi},~\ref{lemma:stabs},~\ref{lemma:stapp}
  can be used to discharge their syntactic counterpart
  (e.g. Lemma~\ref{lemma:stabs} for case \rref{T-Abs}). The remaining
  cases not covered by the lemmas are similar to the ones already
  shown or simpler and therefore omitted from the text.

\end{proof}

\scw{Spell this out in a little more detail.}

\begin{proof}
  Immediate from Theorem~\ref{theorem:soundness} and the $[[Void]]$ case of Lemma~\ref{lemma:interpinv}.
\end{proof}

@@

\Rref{P-AbsEta} effectively adds $\eta$ laws for functions to our
equational theory since coherence, the untyped relation used for type
conversion, is built on top of parallel reduction. We can recover the
same confluence result about parallel reduction using the standard
techniques from \citet{barendregt:lambda-calculi-with-types,
takahashi-parallel-reduction}, though specifically for the de Bruijn
representation, we need the following anti-renaming lemma about
parallel reduction.
\begin{lemma}[Par anti-renaming]
  \label{lemma:parantirenaming} If $[[a < xi > => b0]]$, then there
exists some $[[b]]$ such that $[[b < xi > = b0]]$ and $[[a => b]]$.
\end{lemma}

@@

The predicates $[[wne a]]$ and $[[wn
a]]$ describe terms that can evaluate into $\beta$-neutral or
$\beta$-normal form through parallel reduction and are defined as
follows.
\[ [[wne a]] \iff \exists [[e]], [[a =>+ e]] \]
\[ [[wn a]] \iff \exists [[f]], [[a =>+ f]] \]
We can read $[[wne a]]$ and $[[wn a]]$ as weakly normalizing to a neutral or a normal form respectively.

@@

Before we can prove the fundamental theorem and derive the
normalization property as its corollary, we need to additionally
formulate and prove the adequacy property.

>>>>

% \scw{Need to explain the notations in the text. What are these operations?
% What do readers need to understand about them? Help me understand
% the figure} \yl{resolved}
\scw{As this is a tutorial paper, you'll need to explain more about how de Bruijn
indices work.}
% We omit most of the
% definitions of renaming and substitution and only show the definition
% of a few representative cases of substitution.
\begin{figure}[ht]
  \begin{equation*}
    \begin{split}
      \begin{array}{lll}
        % \mathit{IdentityRen} \\
        % [[id i]] & := & [[i]] \\ \\
        \mathit{IdentityTm} \\
        [[idtm n]] & := & [[var n]] \\ \\
        % \mathit{ConsRen ([[xi .: n]])} \\
        % ([[xi .: n]])([[0]]) & := & [[n]] \\
        % ([[xi .: n]])([[Suc i]]) & := & [[xi i]] \\ \\
        \mathit{ConsSubst ([[rho .: a]])} \\
        ([[rho .: a]])([[0]]) & := & [[a]] \\
        ([[rho .: a]])([[Suc n]]) & := & [[rho n]] \\ \\
        \mathit{Curried\ Add} \\
        [[up n m]] & := & \ottkw{v}_{[[n]] + [[m]]}
      \end{array}
    \end{split}
    \qquad \qquad
    \begin{split}
      \begin{array}{lll}
        % \mathit{UpRen} \\
        % [[( up xi ) 0]] & := & [[0]] \\
        % [[( up xi ) Suc i]] & := & [[ Suc ren xi i ]] \\ \\
        % \mathit{Renaming ([[a < xi >]])} \\
        % [[(Pi A B) < xi >]] & := & [[Pi A < xi > B < up xi >]] \\
        % [[(a b) < xi >]] & := & [[a < xi > (# b < xi > #)]] \\
        % \ldots \\ \\
        \mathit{Lifting} \\
        [[( up rho ) 0]] & := & [[var 0]] \\
        [[( up rho ) Suc n]] & := & [[ rho n < up 1 > ]] \\ \\
        \mathit{Substitution ([[ a { rho }  ]])} \\
        [[var n { rho }  ]] & := &  [[rho n]] \\
        [[(Pi A B) { rho }]] & := & [[Pi A { rho } B { up rho }]] \\
        [[(a b) { rho }]] & := & [[a { rho } (# b { rho } #)]] \\
        [[(\ a) { rho }]] & := & [[\ (a { up rho })]] \\
        \ldots
      \end{array}
    \end{split}
  \end{equation*}
  \caption{Auxiliary Functions over Syntax}
  \label{fig:auxdef}
\end{figure}

@@

\section{Semantic Typing and Consistency}
\label{sec:logrelproof}
\yl{todo: rename all variables from i,j to n}
\begin{figure}[h]
\[
\begin{array}{lcl}

@@

The necessity of semantic typing as an extra layer of definition on
top of the logic relation can be understood in simply typed
languages~\citep{skorstengaard2019introduction, harpertait,
  pierce2002types}. In our setting, attempting to show that $[[empty
|- a : A]]$ implies $[[InterpR i A S]]$ and $[[a in S]]$ will fail in
\rref{T-Abs}, where the induction hypothesis is not helpful since the body
of the lambda term is typed under a non-empty context. Through
the definition of semantic typing, we can state a strengthened
property that is actually provable.

@@

The $[[rho |= G]]$ notation denotes the semantic well-formedness of
a substitution $[[rho]]$ with respect to a context
$[[G]]$. Intuitively, for variable $[[i]]$ in the context (thus the
restriction $[[i < | G |]]$), $[[rho i]]$ is a term that inhabits all
possible interpretations of the type $[[G i]]$ from the
context.
Since we are in a dependently typed system, there
may be variables in $[[G i]]$, too. The type $[[(G i < up Suc i > ) {
  rho }]]$ applies $[[rho]]$ to replace those variables. The shifting
operation $[[up Suc i]]$ can be viewed as an artifact of the de Bruijn
representation to lift the type $[[G i]]$ to the right scope before
substitution can be carried out.
Finally, despite the for all quantification over the interpreted set
$[[S]]$ and the universe level $[[j]]$ in the definition of $[[rho |=
G]]$, we only need show that there exists some $[[j]]$ and $[[S]]$
such that $[[InterpR j (G i < up Suc i > ) { rho } S]]$ and $[[rho i
in S]]$. This idea is encapsulated in the second of the two structural
law about $[[rho |= G]]$.
\begin{lemma}[Well-formed $[[rho]]$ empty\footnote{\dotv{soundness.v}{$\gamma$\_ok\_nil}}]
  \label{lemma:rhowfempty}
  $[[rho |= G]]$ whenever $[[G]]$ is the empty context.
\end{lemma}
\begin{lemma}[Well-formed $[[rho]]$ cons\footnote{\dotv{soundness.v}{$\gamma$\_ok\_cons}}]
  If $[[InterpR i A S]]$, $[[a in S]]$, and $[[rho |= G]]$, then
  $[[rho .: a |= G ++ A]]$.
\end{lemma}

We refer to $[[G |= a : A]]$ as semantic well-typedness. Its definition says the term $[[a]]$ can be
semantically typed $[[A]]$ under the context $[[G]]$ if for all
substitutions $[[rho]]$ such that $[[rho |= G]]$, the type $[[A { rho }]]$ can be interpreted
as the set $[[S]]$, and $[[a { rho } in S]]$. Our definition of
semantic well-typedness is standard, though dependent types add a
small twist that we apply the $[[rho]]$ to $[[A]]$ and require that
$[[A { rho }]]$ has some interpretation, which is not always the case
for arbitrary types.

Analogous to $[[|- G]]$, the syntactic well-formedness of the typing
context, $[[|= G]]$ defines when a context is semantically
well-formed. In the
definition of $[[|= G]]$, we use the notation $[[drop i G]]$ to denote
the typing context obtained by dropping the last $[[i]]$ elements of
$[[G]]$. When $[[G]]$ has less than $[[i]]$ elements, $[[drop i G]]$
returns the empty list. Intuitively, the definition of $[[|= G]]$
simply says all types in the context must correspond to some set
according to the logical relation. The truncation to the context
$[[G]]$ is due to the different scoping of the types in the
context. An alternative way of defining $[[|= G]]$ would look more
similar to $[[|- G]]$, though we find it easier to define $[[|= G]]$
through for all quantification of $[[i]]$ and recover the structural
rules as lemmas.
\begin{lemma}[Semantic context well-formedness cons]
  \label{lemma:semwffcons}
  If $[[|= G]]$ and $[[G |= A : Set i]]$, then $[[|= G ++ A]]$.
\end{lemma}
\begin{lemma}[Semantic context well-formedness empty]
  \label{lemma:semwffempty}
  $[[|= G]]$ holds when $[[G]]$ is empty.
\end{lemma}

The following lemma makes the statement $[[G |= A : Set i]]$ easier to
work with.
\begin{lemma}[Set Inversion\footnote{\dotv{soundness.v}{SemWt\_Univ}}]
  \label{lemma:setinv}
  The following two statements are equivalent:
  \begin{itemize}
  \item $[[G |= A : Set i]]$
  \item $\forall$ $[[rho]]$, if $[[rho |= G]]$, then there exists
    $[[S]]$ such that $[[InterpR i (A {rho}) S]]$
  \end{itemize}
\end{lemma}
\begin{proof}
  The forward direction is immediate by
  Lemma~\ref{lemma:interpinv}. We now consider the backward direction
  and show that $[[G |= A : Set i]]$ given the second bullet.

@@

% The semantic context well-formedness judgment ($[[|= G]]$), unlike its syntactic
% counterpart $[[|- G]]$, is defined through a $\forall$ quantified statement rather
% than inductively over the context. \scw{why is that the case?} It is easy to recover the same
% structural rules:
% \scw{point out that this lemma is the semantic analogue to Ctx-Cons. What about
% Ctx-empty?}
% \begin{proof}
%   By the definition of semantic context well-formedness, the goal is
%   to show that given $[[i < | G ++ A |]] = [[Suc |G|]]$, $[[drop Suc i   (G ++ A) |= (G ++ A) i : Set j]]$. The statement can be easily
%   proven by case analysis on whether $[[i]]$ is zero.
% \end{proof}

Next, we show some non-trivial cases of the fundamental theorem as
top-level lemmas.
\begin{lemma}[ST-Var]
  \label{lemma:stvar}
  If $[[|= G]]$ and $[[i < |G|]]$, then $[[G |= var i : G i < up Suc i  >]]$.
\end{lemma}
\begin{proof}
  Suppose $[[rho |= G]]$. By the definition of semantic typing, we
need to show that there exists some $[[j]]$ and $[[S]]$ such that
  \begin{itemize}
  \item $[[InterpR j G i < up Suc i > { rho } S]]$
  \item $[[rho i in S]]$
  \end{itemize}
  From the definition of $[[|= G]]$, we know that there exists some
  $[[j]]$ such that $[[drop Suc i G |= G i : Set j]]$. This is not
  immediately useful since $[[rho |= G]]$, not the truncated
  context. We claim, without giving the proof, that from $[[drop Suc i
  G |= G i  : Set j]]$, we can derive $[[G |= G i {up Suc i} : Set
  j]]$. A rigorous justification would require us to formally state
  and show renaming/weakening lemma for semantic typing. We instead
  provide an informal justification since the proof of renaming
  is uninteresting and only requires unfolding definitions, but its
  statement introduces unnecessary line noise related to shifting.

  Intuitively, a substitution $[[rho]]$
  that closes over a context should always induce a truncated
  substituion for each truncated context. As a result, if
  some term $[[b]]$ is semantically well-typed under some truncated
  context $[[G]]$, then it must also be semantically well-typed under
  the context $[[G]]$ up to shifting.

  Knowing that $[[G |= G i {up Suc i} : Set j]]$ holds, we can then
  apply Lemma~\ref{lemma:setinv} to show that
  $[[InterpR j G i < up Suc i > { rho } S]]$. It now
  suffices to show $[[rho i in S]]$, but that is immediate from the
  definition of $[[rho |= G]]$.
\end{proof}

% Next, we show some non-trivial cases of the fundamental theorem as
% top-level lemmas.
% We first formulate the definition of valid renamings and prove that
% semantic typing satisfies renaming so we can weaken the context when
% reasoning about the variable case of the fundamental lemma
% (Lemma~\ref{lemma:stvar}). Intuitively, given a valuation $[[rho |= G ++ D]]$, it is easy to show that we can extract some valuation
% $[[rho0]]$ such that $[[rho0 |= G]]$, where $[[rho0]]$ is obtained by
% ``truncating'' $[[rho]]$. As a result, if we know that $[[G |= a :A]]$, then we can conclude that $[[G ++ D |= a0 : A0 ]]$, where
% $[[a0]]$ and $[[A0]]$ are obtained by shifting $[[a]]$ and $[[A]]$
% after weakening the context; this implication holds because $[[rho |=G ++ D]]$ induces a context $[[rho0]]$ such that $[[rho0 |= G]]$ so we
% can make use of the premise $[[G |= a : A]]$ to derive what we need
% for the conclusion. We recommend the readers to skip the proofs of
% Lemmas~\ref{lemma:validtruncate} through \ref{lemma:semrenaming}
% during the first read as long as they have an intuitive understanding
% of what the renaming property is meant to capture.

% \scw{Make this a definition? Would it be useful to create notation
%   for the relation, such as $[[xi]]:[[G]]\Rightarrow[[D]]$? }
% We say that $[[xi]]$ is
% valid from the context $[[G]]$ to the context $[[D]]$ if the
% following condition holds.
% \[ \forall i \text{, if } i < | [[G]] |\text{, then }[[ ren xi i < |D| ]] \text{ and }
%   [[D ren xi  i < up Suc ren xi i > = G i < up Suc i > < xi >]] \]

% \begin{lemma}[Truncate is valid]
%   \label{lemma:validtruncate}
%   If $[[i < |G|]]$, then $[[up i]]$ is a valid renaming from $[[drop i   G]]$ to $[[G]]$.
% \end{lemma}
% \begin{proof}
%   By the definition of a valid renaming, we must show that given $[[j   < |drop i G|]] = [[|G|]] - [[i]]$, then the following conditions hold:
%   \begin{itemize}
%   \item $[[ren up i j]] = i + j  < [[| G | ]]$
%   \item $[[G ren up i j < up Suc ren up i j > = drop i G j < up Suc j > < up i >]]$
%   \end{itemize}
%   The first bullet point is immediate from the fact that $[[j]] <
%   [[|G|]] - [[i]]$. The second bullet follows from unfolding the
%   definitions, and the fact that $[[drop i G j]] = [[G]](i + j)$ when
%   $[[i]] + [[j]] < [[|G|]]$.
% \end{proof}

% \begin{lemma}[Renaming for $[[rho |= G]]$]
%   \label{lemma:renamingval}
%   If $[[rho |= D]]$ and $[[xi]]$ is a valid renaming from $[[G]]$ to
%   $[[D]]$, then we have $[[rho0 |= G]]$ where $[[rho0 i = rho ren xi i]]$.
% \scw{could you use $[[rho]]\circ[[xi]]$ for $[[rho0]]$?}
% \end{lemma}
% \begin{proof}
%   Unfolding the definition of $[[rho0 |= G]]$, the goal is to show
%   that for all $i,j,$ and $S$, if $[[i < |G|]]$ and $[[InterpR j (G i   < up Suc i > ) { rho0 } S ]] \text{, then } [[rho0 i]] = [[rho ren xi i in S]]$.

%   By the definition of $[[rho0]]$ and the validity of $[[xi]]$, we
% have $[[G i < up Suc i > { rho0 }]] = [[G i < up Suc i > < xi > { rho } ]] = [[D ren xi i < up Suc ren xi i > { rho }]]$. From $[[rho |= D]]$ and $[[ren xi i < | D |]]$, we have $[[InterpR j D ren xi i < up Suc ren xi i > { rho } S]]$ and $[[rho ren xi i in S]]$.
% \end{proof}

% \begin{lemma}[Renaming for $[[G |= a : A]]$]
%   \label{lemma:semrenaming}
%   If $[[G |= a : A]]$ and $[[xi]]$ is a valid renaming from $[[G]]$ to
%   $[[D]]$, then $[[D |= a < xi > : A < xi > ]]$.
% \end{lemma}
% \begin{proof}
%   Immediate from the definition of semantic typing and Lemma~\ref{lemma:renamingval}.
% \end{proof}

% \scw{Need a transition here that you are starting to explain the semantic typing
% rules.}

@@

  By Lemma~\ref{lemma:setinv} and the premise $[[G |= Pi A B : Set
  i]]$, there exists some set $[[S]]$ such that
  $[[InterpR i Pi A {rho} B {up rho} S]]$. It now suffices to show that
  $[[\A{rho} b{up rho} in S
  ]]$. By Lemma~\ref{lemma:piinvalt}, there exists some $[[S0]]$ such
  that all following conditions hold:
  \begin{itemize}
  \item $[[InterpR i A{rho} S0]]$
  \item $[[forall a, (# a in S0 implies (# exists S1 , InterpR i B
    {rho .: a}
    S1 #) #)]]$
  \item $[[S = { b | forall a, (# a in S0 , forall
      S1, (# InterpR i B {rho .: a} S1,  b a in S1 #) #) }]]$
  \end{itemize}
  To show that $[[\A{rho} b{up rho} in S]]$, we need to prove
  that given $[[a in S0]]$,
  $[[Interp I i B {rho .: a} S1]]$, we have  $[[( \A{rho} b{up rho} )
  a in S1]]$.
  By Lemma~\ref{lemma:logrelbackclos}, the set $[[S1]]$ is closed
  under expansion. Since $[[( \A{rho} b{up rho} )
  a => b {up rho} {a}]] = [[b {rho .: a}]]$, it suffices to show that
  $[[b {rho .: a} in S1]]$, which is immediate from $[[G ++ A |= b :
  B]]$ and the fact that the logical relation is deterministic and
  cumulative (Lemma~\ref{lemma:logreldeterhet}).
\end{proof}

@@

\begin{theorem}[The Fundamental Theorem\footnote{\dotv{soundness.v}{soundness}}]
  \label{theorem:soundness}\leavevmode
  \begin{itemize}
  \item If $[[G |- a : A]]$, then $[[G |= a : A]]$.
  \item If $[[|- G]]$, then $[[|= G]]$.
  \end{itemize}
\end{theorem}
\begin{proof}
  Proof by mutual induction over the derivation of $[[G |- a :
  A]]$ and $[[|- G]]$.   The cases related to context well-formedness immediately follows
  from Lemma~\ref{lemma:semwffcons}.
Lemmas~\ref{lemma:stvar},~\ref{lemma:stset},~\ref{lemma:stpi},~\ref{lemma:stabs},~\ref{lemma:stapp}
  can be used to discharge their syntactic counterpart
  (e.g. Lemma~\ref{lemma:stabs} for case \rref{T-Abs}). The remaining
  cases not covered by the lemmas are similar to the ones already
  shown or simpler and therefore omitted from the text.
\end{proof}

Recall the logical consistency property
(Theorem~\ref{theorem:consistency}), which states that the judgment
$[[empty |- a : Void ]]$ is not derivable. We now give a proof of the
property using the fundamental lemma.

\begin{proof}
  Suppose $[[empty |- a : Void]]$ is derivable, then by the
  fundamental lemma, we have $[[empty |= a : Void]]$, which states
  that for all $[[rho |= empty]]$, and for all $[[j]]$, $[[S]]$ such
  that $[[InterpR j Void S]]$, we have $[[a {rho} in S]]$. By
  Lemma~\ref{lemma:rhowfempty}, any $[[rho]]$ we pick trivially
  satisfies $[[rho |= G]]$. For convenience, we pick $[[rho]]$ as
  $[[idtm]]$ (though any $[[rho]]$ would work since
  $[[empty |- a : Void]]$ ensures there is no free variable in
  $[[a]]$), we have $[[a {idtm}]] = [[a in S]]$. By the $[[Void]]$
  case of the inversion property (Lemma~\ref{lemma:interpinv}), we
  know that $[[S]]$ must be the empty set and therefore contradicts
  the assupmtion that $[[a in S]]$. Done.
\end{proof}

@@

\Rref{P-AbsEta} effectively adds the $\eta$ law for functions to our
equational theory since convertibility ($[[a <=> b]]$), the untyped relation used for type
conversion, is built on top of parallel reduction. We can recover the
same confluence result about parallel reduction using the standard
techniques from \citet{barendregt:lambda-calculi-with-types,
takahashi-parallel-reduction}, though specifically for the de Bruijn
representation, we need the following anti-renaming lemma about
parallel reduction.
\begin{lemma}[Par anti-renaming]
  \label{lemma:parantirenaming} If $[[a < xi > => b0]]$, then there
exists some $[[b]]$ such that $[[b < xi > = b0]]$ and $[[a => b]]$.
\end{lemma}
Both the confluence and anti-renaming properties require structural
induction over the size of terms to obtain a useful induction
hypotehsis in the $\eta$ case.

@@

The predicates $[[wne a]]$ and $[[wn
a]]$ describe terms that can evaluate into $\beta$-neutral or
$\beta$-normal form through parallel reduction and are defined as
follows.
\[ [[wne a]] \iff \exists [[e]], [[a =>+ e]] \]
\[ [[wn a]] \iff \exists [[f]], [[a =>+ f]] \]
We can read $[[wne a]]$ and $[[wn a]]$ as the term $[[a]]$ weakly normalizes to a neutral or a normal form respectively.

@@

Before we can prove the fundamental theorem and derive the
normalization property as its corollary, we need to additionally
formulate and prove the adequacy properties.

Let us first start with some auxiliary lemmas and definitions.
\begin{lemma}[Ext Wn]
  \label{lemma:extwn}
  If $[[wn (a var i)]]$, then $[[wn a]]$.
\end{lemma}
\begin{proof}
  By induction over the length of the reduction sequence in $[[wn (a
  var i)]]$. The conclusion follows from Lemma~\ref{lemma:parantirenaming} and
  \ref{lemma:parnenf}.
\end{proof}
